---
title: "STAT394 / Milestone 4"
author: "Group 7"
date: '12-08-2022'
header-includes:
    - \usepackage{newpxtext,eulerpx}
    - \usepackage{bm,bbm}
    - \usepackage{palatino}
    - \renewcommand{\familydefault}{\sfdefault} % sans serif
    - \fontfamily{ppl}\selectfont
    - \usepackage{float}
    - \floatplacement{figure}{H}
    - \usepackage{booktabs}
output: bookdown::pdf_document2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE}
library(readxl)
library(e1071)
library(ggplot2)
library(ggcorrplot)
library(GGally)
library(ggExtra)
library(car)
library(factoextra)
library(cluster)
library(ggbiplot)
library(rgl)
```

```{r, echo=FALSE}
taslakes.data <- data.frame(read_excel("../../DATA/EXCEL/Tasmania_lakes_recoded.xlsx", sheet = 2))
```

# Consolidated Exploratory Data Analysis

## Overview: Tasmanian Lakes

The dataset features 63 variables (environmental indicators, spatial data, etc.) for 50 Tasmanian lakes. Understanding how natural environmental factors (rock type, precipitation, etc.) may impact lake quality indicators (pH, temperature, etc.) is of particular interest. This is because Tasmania features a strong east-west binary division in many natural characteristics, including rock type and precipitation. Understanding the extent to which natural drivers control a baseline for lake quality - and how these baseline conditions may differ between lakes - is especially important when setting targets for lake health restoration. For example, if a lake is naturally very acidic, then resources should not be allocated towards raising pH back to neutral. 
Eleven key variables were identified as particularly important in the initial EDA: lake elevation, lake depth, lake pH, water temperature, precipitation, lake turbidity, lake area, the presence of glacial sediment, the presence of scrub, the presence of felsic rock and the presence of dolerite. The latter four variables are categorical, and the final two are proxies for two distinct rock classifications.

```{r echo=FALSE, include=FALSE}
GlaSed <- factor(taslakes.data$GlaSed)
Scrub <- factor(taslakes.data$Scrub)
Felsic <- factor(taslakes.data$Felsic)
Dolerite <- factor(taslakes.data$Dolerite)
Elev <- taslakes.data$Elev
Depth <- taslakes.data$Depth
pH <- taslakes.data$pH
Twater <- taslakes.data$Twater
Precip <- taslakes.data$Precip
Turbid <- taslakes.data$Turbid
Lake_Area <- taslakes.data$Lake_Area
Lake <- taslakes.data$Lake

taslakes.df <- data.frame(taslakes.data$Lake, taslakes.data$Elev, taslakes.data$Depth, taslakes.data$pH, taslakes.data$Twater, taslakes.data$Precip, taslakes.data$Turbid, GlaSed, Scrub, Felsic, Dolerite, taslakes.data$Lake_Area)
```

## Interaction Table: Rock Types

An important interaction is the exclusivity of the different rock types ('Felsic' and 'Dolerite'). Table&nbsp;\ref{tab:Table2}, below, demonstrates that none of the sampled lakes have both rock types, and many have neither. This motivates further study to explore if the distinct and exclusive nature of these rock types may influence variations in environmental factors at the lakes. 

``` {r Table2, echo=FALSE, results = 'asis'}

levels(taslakes.df$Felsic) <- c("Not Felsic", "Felsic", "Felsic")
levels(taslakes.df$Dolerite) <- c("Not Dolerite", "Dolerite")

rocktype.int <- table(interaction(taslakes.df$Felsic, taslakes.df$Dolerite))

Interaction <- c("Not Felsic + Not Dolerite", "Felsic + Not Dolerite", "Not Felsic + Dolerite", "Felsic + Dolerite") 
Frequency <- c(24, 4, 22, 0)

rocktype.int.table <- data.frame(Interaction, Frequency)

knitr::kable(rocktype.int.table,
             caption="Interaction of Rock Types 'Felsic' and 'Dolerite'", 
             format="latex", booktabs=TRUE, position = "!ht")
```

## Descriptive Summary Statistics

```{r echo=FALSE,include=TRUE}

range2 <- function(x) {
  range <- max(x) - min(x)
  return(range)
}

summary <- function(x){
  return(c(
    length(x),
    min(x),
    quantile(x, .25),
    median(x),
    quantile(x, .75),
    max(x),
    IQR(x),
    range2(x),
    mean(x),
    sd(x),
    skewness(x),
    kurtosis(x))
  )
}
```

Table&nbsp;\ref{tab:Table3}, below, presents a quantitative summary of the descriptive statistics for each of the seven numeric variables. 

```{r Table3, echo=FALSE, results = 'asis'}

taslakes.df.num <- taslakes.df[, -c(1,(8:11))]

taslakes.summarytable <- apply(taslakes.df.num, summary, MARGIN=2)

taslakes.summarytable <- rbind(formatC(as.numeric(taslakes.summarytable[1,]),format="f",digits=0),
                  formatC(as.numeric(taslakes.summarytable[2,]),format="f", digits=2),
                  formatC(as.numeric(taslakes.summarytable[3,]),format="f", digits=2),
                  formatC(as.numeric(taslakes.summarytable[4,]),format="f", digits=2),
                  formatC(as.numeric(taslakes.summarytable[5,]),format="f", digits=2),
                  formatC(as.numeric(taslakes.summarytable[6,]),format="f", digits=2),
                  formatC(as.numeric(taslakes.summarytable[7,]),format="f", digits=2),
                  formatC(as.numeric(taslakes.summarytable[8,]),format="f", digits=2),
                  formatC(as.numeric(taslakes.summarytable[9,]),format="f", digits=2),
                  formatC(as.numeric(taslakes.summarytable[10,]),format="f", digits=2),
                  formatC(as.numeric(taslakes.summarytable[11,]),format="f", digits=2),
                  formatC(as.numeric(taslakes.summarytable[12,]),format="f", digits=2))


colnames(taslakes.summarytable) <- c("Elevation", "Depth", "pH", "Water Temp", "Precipitation", "Turbidity", "Lake Area")
rownames(taslakes.summarytable) <- c("Sample Size", "Minimum", "Lower Quartile", "Median", "Upper Quartile", "Maximum", "IQR", "Range", "Mean", "Standard Deviation", "Skewness", "Kurtosis")

knitr::kable(taslakes.summarytable,
             caption="Quantitative summary of numeric Tasmania lakes data", 
             format="latex", booktabs=TRUE, position = "h!")

```

Table&nbsp;\ref{tab:Table3} suggests that the different numeric variables may present different distributions. Turbidity, Depth and Lake Area seem the most asymmetric, with the highest 'skewness'. For each of these three variables, means are greater than medians, suggesting that the assymetry follows a right-skew whereby a smaller number of highly turbid, deep and large lakes are in the sample. 

Turbidity also has a very high 'kurtosis', suggesting that its distribution is very tail-heavy. Precipitation and pH appear to be distributed the most symmetrically, with the smallest 'skewness' and means that more closely approximate their medians.

The causes in which assumptions of Normality are challenged (i.e. variables with skewed distributes) must be considered when fitting models later in the report. 

## Visualing the Correlation Matrix

Exploring correlations between different numeric variables may help inform interesting relationships to investigate further. Figure&nbsp;\@ref(fig:Figure1), below, provides a visual interpretation of the correlation matrix.

```{r Figure1, fig.cap="Plot of the numeric Tasmania lakes data correlation matrix", fig.pos="center",echo=FALSE}

taslakes.cor.matrix <- matrix(c(cor(taslakes.df.num)), nrow=7)

colnames(taslakes.cor.matrix) <- c("Elevation", "Depth", "pH", "Water Temp", "Precipitation", "Turbidity", "Lake Area")
rownames(taslakes.cor.matrix) <- c("Elevation", "Depth", "pH", "Water Temp", "Precipitation", "Turbidity", "Lake Area")

ggcorrplot(cor(taslakes.cor.matrix),
           hc.order = TRUE,
           type = "lower", 
           legend.title ="Correlation", 
           outline.color = "black",
           ggtheme = theme_bw())
```

Some of the correlations presented in Figure&nbsp;\@ref(fig:Figure1) are expected, and some may be of interest for further investigation. For example, Water Temperature appears to be strongly negatively correlated with Elevation, and Depth and Lake Area appear positively correlated. This implies that water temperature decreases as elevation increases and lakes with larger areas are usually deeper, both expected relationships.

Of interest, Elevation appears to have a strong positive correlation with pH. This implies that as the elevation of lakes increases, pH increases and water temperature decreases. pH also appears to have a strong negative correlation with Precipitation. What may be driving these correlations are less clear, and may be appropriate subjects for quantitative study in this report.

## Pairs Plot

Figure&nbsp;\@ref(fig:Figure2), below, provides the pairs plot.

```{r Figure2, fig.cap="Pairs plot for numeric Tasmania lakes data", fig.pos="center",echo=FALSE}

ggpairs(taslakes.df.num)
```

Figure&nbsp;\@ref(fig:Figure2) provides more insight into the distributions of each variable, after preliminary interpretations were made from the summary statistics. Strong asymmetry amongst is shown in several variables, including Turbidity, Depth and Lake Area. Though all other variables are less skewed, none of the seven appear to follow a Gaussian distribution.

In particular, Precipitation and pH appear to follow a bimodal distribution. Both  variables should be investigated further, as their distributions may inform the presence of distinct sub-groups (i.e. sub-populations) of lakes that have been combined together in the sample. Referring back to Figure&nbsp;\@ref(fig:Figure1), Precipitation and pH exhibited a moderate-to-strong negative correlation. Given that relationship, the similarity in their distributions m2ay imply that one variable is driving the other. This is explored further in Figure&nbsp;\@ref(fig:Figure3).

## Investigating Bimodal Distributions: pH

Figure&nbsp;\@ref(fig:Figure3) and Figure&nbsp;\@ref(fig:Figure4) examine how the bimodal distribution of pH may be explained by the presence of absence of Dolerite rock and Elevation, respectively. 

### Multiple Density Plot: pH, Dolerite

Background research on the topic has suggested unique combinations of minerals are released when different rock type erode (e.g. Dolerite vs no Dolerite). These minerals then wash into lakes and possibly affect pH, such that rock type may drive lake pH.

```{r Figure3, fig.cap="Multiple density plot of pH for the presence or absence of Dolerite", fig.pos="center",echo=FALSE}

taslakes.df$Dolerite <- factor(taslakes.df$Dolerite)

ggplot(taslakes.df, aes(x=pH, color=Dolerite, fill=Dolerite)) + 
  geom_density(alpha=.7)
```

Figure&nbsp;\@ref(fig:Figure3) provides a strong indication that the two groups of lakes (lakes with Dolerite rock, lakes without Dolerite rock) have different mean pH. Specifically, lakes with Dolerite rock appear have have a higher pH with less variance than lakes without Dolerite. This will be examined further with inferential statistics.


### Scatterplot: Elevation, pH, Dolerite

```{r Figure4, fig.cap="Scatterplot (with rugplot) plot of pH against Elevation, controlled for the presence or absence of Dolerite", fig.pos="center",echo=FALSE}

ggplot(taslakes.df, aes(x=pH, y=Elev, color=Dolerite)) +
  geom_text(label=Lake) +
  geom_smooth(method=lm, se=FALSE) +
  geom_rug()

```

Figure&nbsp;\@ref(fig:Figure4) suggests that lakes without Dolerite rock may have systematically lower pH values and be located at lower Elevations than lakes with Dolerite rock. Pending inferential analyses, this may imply that Elevation differences also play a role in producing the bimodal pH distribution. For lakes without Dolerite rock, pH appears to increase with Elevation, while for lakes with Dolerite rock pH may decrease with Elevation, though the trend is less discernable.


# Hypothesis Testing


## Testing Assumptions (T Test and ANCOVA)

The following tests and checks are used to evaluate to extent to which assumptions that apply to both t tests and ANCOVA are met. If key assumptions (e.g. Normality) are not upheld, then a non-parametric (Kruskal-Wallis) test will be used.

### Linearity

The assumption of linearity between the response (pH) and the covariate (Elevation) can be assessed and upheld based on Figure&nbsp;\@ref(fig:Figure4). For both groups of lakes (with and without Dolerite), there appears to be a linear relationship between pH and Elevation.

### Homogeneity of Regression Slopes

```{r echo=TRUE}

interaction <- aov(pH ~ Dolerite*Elev)
rm(summary)
summary(interaction)
```

The assumption of homogeneity of regressions slopes is upheld as the p-value of $0.2033$ suggests that the interaction between the covariate (Elevation) and group (Dolerite) is not significant. 

### Normality of Residuals

```{r Figure5, fig.cap="Normal QQ Plot for pH", fig.pos="center",echo=FALSE}

qqnorm(pH, frame = FALSE)
qqline(pH, col = "dark blue", lwd = 3)
```

```{r Figure6, fig.cap="Normal QQ Plot for Elevation", fig.pos="center",echo=FALSE}

qqnorm(Elev, frame = FALSE)
qqline(Elev, col = "dark red", lwd = 3)
```

The assumption of Normality is challenged by deviates straying from Normality for both variables at extreme observations (pH in Figure&nbsp;\@ref(fig:Figure5), Elevation in Figure&nbsp;\@ref(fig:Figure6)). The results of the t test and ANCOVA must therefore be treated with some caution, but the analysis will proceed.

### Homogeneity of Variances

```{r Figure7, fig.cap="Boxplots to demonstrate Normality and variance across the with and without Dolerite lake groups", fig.pos="center",echo=FALSE}

ggplot(taslakes.df, aes(x=Dolerite, y=pH, fill=Dolerite)) + 
  geom_boxplot()
```

Figure&nbsp;\@ref(fig:Figure7) suggests what was also demonstrated in Figure&nbsp;\@ref(fig:Figure3): the two groups (lakes with and without Dolerite) do not have equal variance, and may be slightly right skewed. The lakes without Dolerite have a much larger spread. 

```{r echo=TRUE}

leveneTest(pH ~ Dolerite)
leveneTest(Elev ~ Dolerite)
```

The assumption of equal variances appears not to be upheld through Levene's Tests of pH and Elevation across lakes with and without lake groups, as in both cases the p value $< 0.05$. Welch's t-test will be used, such that equal variances are not assumed. The results of the ANCOVA must therefore be treated with some caution, but the analysis will proceed. 

## T Test 

$H_0$: true difference in means is equal to 0, 
$H_1$: true difference in means is not equal to 0

```{r echo=TRUE}

pH.Dol.ttest <- t.test(pH ~ Dolerite, var.eq=FALSE)
pH.Dol.ttest
```

The t test reports a p-value of $2.506*10^{-14}$ based on a test statistic of $-11.805$ on $38.281$ degrees of freedom. There is sufficient evidence to reject $H_0$ and conclude that the difference in mean pH between lakes with and without Dolerite is not 0.       

## Kruskal-Wallis Test

$H_0$: means of treatment groups do not differ,
$H_1$: means of treatment groups differ

The Kruskal-Wallis test will be used as a non-parametric alternative to the t test, as Figure&nbsp;\@ref(fig:Figure6) and Figure&nbsp;\@ref(fig:Figure7) suggest that the assumption of Normality may not be upheld.

```{r echo=TRUE}

kruskal.test(pH ~ Dolerite)
```

Agreeing with the t test, the Kruskal-Wallis test provides enough evidence against $H_0$ as the p-value is $2.773*10^{-9}$. In both parametric and non parametric analyses, pH appears to systematically differ between lakes with and without Dolerite.


## ANCOVA

The ANCOVA adjusts for the effect of Elevation as a possible nuisance variable obscuring the relationship between pH and presence/absence of Dolerite identified in the t test and Kruskal-Wallis test. The parametric nature of the ANCOVA test means that caution must be exercised in the interpretation, as there was evidence against Normality in the response and covariate. 

$H_0$: true difference in means is equal to 0, controlling for the covariate,
$H_1$: true difference in means is not equal to 0, controlling for the covariate

```{r echo=TRUE}

tas.ancova <- aov(pH ~ Elev + Dolerite)
summary(tas.ancova)
```

The ANCOVA reports a p-value of $2.17*10^{-10}$ based on a test (F) statistic of $64.75$ for the difference in mean pH between the two groups after adjusting for Elevation. There is sufficient evidence to reject $H_0$ and conclude that the difference in mean pH between lakes with and without Dolerite is not 0 after adjusting for Elevation. 

# Principal Components Analysis and Outliers

## Standarding and Stretching Numeric Variables

A principal components analysis can be performed on the data after standardising and stretching each numeric variables to [0,1].


```{r echo=FALSE}
range01 <- function(x){(x-min(x))/(max(x)-min(x))}
stretch <- function(x) { return(2*range01(x)-1)
}
std <- function(x){(x-mean(x))/sd(x)}
range01 <- function(x) { minimax <- range(x)
  return((x-minimax[1])/(minimax[2]-minimax[1]))
}
#(R-mean(R))/sd(R)
logTransform <- function(x){
  return(log(x))
}

Elev.stretch <- stretch(Elev)
Depth.stretch <- stretch(Depth)
pH.stretch <- stretch(pH)
Twater.stretch <- stretch(Twater)
Precip.stretch <- stretch(Precip)
Turbid.stretch <- stretch(Turbid)
LakeArea.stretch <- stretch(Lake_Area)

df.stretch <- data.frame(Elev.stretch, Depth.stretch, pH.stretch, Twater.stretch, Precip.stretch, Turbid.stretch, LakeArea.stretch)

summary <- function(x){
  return(c(
    length(x),
    min(x),
    quantile(x, .25),
    median(x),
    quantile(x, .75),
    max(x),
    IQR(x),
    range2(x),
    mean(x),
    sd(x),
    skewness(x),
    kurtosis(x))
  )
}
```

```{r Table4, echo=FALSE, results = 'asis'}

taslakes.summarytable.s <- apply(df.stretch, summary, MARGIN=2)

taslakes.summarytable <- rbind(formatC(as.numeric(taslakes.summarytable.s[1,]),format="f",digits=0),
                  formatC(as.numeric(taslakes.summarytable.s[2,]),format="f", digits=2),
                  formatC(as.numeric(taslakes.summarytable.s[3,]),format="f", digits=2),
                  formatC(as.numeric(taslakes.summarytable.s[4,]),format="f", digits=2),
                  formatC(as.numeric(taslakes.summarytable.s[5,]),format="f", digits=2),
                  formatC(as.numeric(taslakes.summarytable.s[6,]),format="f", digits=2),
                  formatC(as.numeric(taslakes.summarytable.s[7,]),format="f", digits=2),
                  formatC(as.numeric(taslakes.summarytable.s[8,]),format="f", digits=2),
                  formatC(as.numeric(taslakes.summarytable.s[9,]),format="f", digits=2),
                  formatC(as.numeric(taslakes.summarytable.s[10,]),format="f", digits=2),
                  formatC(as.numeric(taslakes.summarytable.s[11,]),format="f", digits=2),
                  formatC(as.numeric(taslakes.summarytable.s[12,]),format="f", digits=2))


colnames(taslakes.summarytable) <- c("Elevation", "Depth", "pH", "Water Temp", "Precipitation", "Turbidity", "Lake Area")
rownames(taslakes.summarytable) <- c("Sample Size", "Minimum", "Lower Quartile", "Median", "Upper Quartile", "Maximum", "IQR", "Range", "Mean", "Standard Deviation", "Skewness", "Kurtosis")

knitr::kable(taslakes.summarytable,
             caption="Quantitative summary of numeric Tasmania lakes data", 
             format="latex", booktabs=TRUE, position = "h!")
```

Table&nbsp;\ref{tab:Table4} shows the result of standardising and rescaling the numeric variables, such that the range is contained to a maximum of 1 and a minimum of 0 in each case.

## Determining Outliers

```{r, echo=FALSE}
df.gg <- df.stretch

df.gg$mahalnobis<- mahalanobis(df.gg, colMeans(df.gg), var(df.gg))
df.gg$pvalue <- pchisq(df.gg$mahalnobis, df=3, lower.tail=FALSE)

df.gg[df.gg$pvalue < 0.001,]

```

```{r, echo=FALSE}
df.gg$outlier <- "no"
df.gg$outlier[c(8,32,40)] <- "yes"
#df.stretch$outlier[c(8)] <- "8"
#df.stretch$outlier[c(11)] <- "11"
#df.stretch$outlier[c(32)] <- "32"
#df.stretch$outlier[c(40)] <- "40"
df.gg$outlier<- as.factor(df.gg$outlier)
```

```{r Figure8, fig.cap="Pairs plot to identify outliers amongst numerical variables", fig.pos="center",echo=FALSE}

ggpairs(taslakes.df.num, aes(colour = df.gg$outlier, alpha = 1) , lower  = list(continuous = "points"),
        upper  = list(continuous = "blank"),
        diag  = list(continuous = "blankDiag"))
```

```{r, echo=FALSE}

for (i in colnames(df.stretch)){
   out <- boxplot.stats(df.stretch[,i])$out
   out_ind <- which(df.stretch[,i] %in% c(out))
   out_ind
   if(length(out_ind) >  0){
     #print(cat(i, ":  ", out_ind, " "))
     print(i)
     print(out_ind)
     print(" ")
   }
   
}
#out

```

Figure&nbsp;\@ref(fig:Figure8) identifies outliers (blue) amongst the numerical variables via the Malanobis distance. Three outliers (blue) were identified via testing the p-values from all 50 Malanobis distances (the 8th, 32nd and 40th lake).

## Principal Components Analysis 

```{r Figure9, fig.cap="Principal components within the numeric Tasmanian lakes variables"}
PCA <- prcomp(df.stretch, center=TRUE, scale=TRUE)
plot(PCA, type="l")
```

```{r, echo=FALSE}
rm(summary)
summary(PCA)
```

```{r Figure10, fig.cap="Biplot of principal components within the numeric Tasmanian lakes variables"}
ggbiplot(PCA, obs.scale = 1, var.scale = .5, alpha = .03) +
  ylim(-2, 2)
```

Figure&nbsp;\@ref(fig:Figure9) and Figure&nbsp;\@ref(fig:Figure10) illustrate how only 29.5% of the variance across the lakes is explained by the first principal component and only 24.2% explained by the second. This is not very descriptive. However, the correlation between Elevation and pH appears again here, as Figure&nbsp;\@ref(fig:Figure10) suggests that lakes with high pH values are similar to lakes at high Elevations.

```{r, echo=FALSE}
principal.component.1 <- PCA$x[,1]
dim(principal.component.1) <- dim(taslakes.df[,1])

principal.component.2 <- PCA$x[,2]
dim(principal.component.2) <- dim(taslakes.df[,1])

principal.component.3 <- PCA$x[,3]
dim(principal.component.3) <- dim(taslakes.df[,1])
```

```{r, echo=FALSE}
rgl_init <- function(new.device = FALSE, bg = "white", width = 640) { if( new.device | rgl.cur() == 0 ) {
    rgl.open()
    par3d(windowRect = 20 + c( 0, 0, width, width ) )
    rgl.bg(color = bg )
  }
  rgl.clear(type = c("shapes", "bboxdeco"))
  rgl.viewpoint(theta = 15, phi = 20, zoom = 0.7)
} 
rgl_init()
```

```{r, echo=FALSE}
rgl_init(new.device = TRUE)
plot3d(principal.component.1, principal.component.2, principal.component.3, "PC1", "PC2", "PC3", type="p", col="orange",size=7, radius=200, lwd=30, box=F)
```

# Preliminary Conclusions

There is preliminary evidence to suggest that rock type around a lake (through the analogue presence/absence of Dolerite) affects the pH of that lake. This is an example of how a natural factor (rock type) influences a key water quality indicator that then impacts ecological systems and drinking-water status of lakes. In lake quality restoration efforts in Tasmania, the acceptable range of pH in lakes with Dolerite rock should be set higher than the range for lakes without Dolerite. Variance across all 50 lakes with respect to the numeric variables cannot be explained by one or two dominant principal components: instead, several different axes contribute less than ~30% variance each and suggest that overall lakes characteristics are determined by a range of components. However, through various plots (including of principal components) the strong linearity between Elevation and pH appears. This is an important relationship to adjust for in the aforementioned rock type / pH analyses. 

